---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loadingData, echo=T}
unzip("activity.zip")
activity_data <- read.csv("activity.csv")
activity_data$date <- as.Date(activity_data$date)
hour <- activity_data$interval %/% 100
minute <- activity_data$interval %% 100
timeOfDay <- paste(hour, minute, sep=":")
activity_data$interval <- as.POSIXct(timeOfDay, format="%H:%M")
```

## What is mean total number of steps taken per day?
```{r meanStepsPerDay, echo = T}
library(dplyr)
total_steps_per_day <- group_by(activity_data, date) %>%
  filter(!is.na(steps)) %>%
  summarise(total=sum(steps))
# histogram of total steps
hist(total_steps_per_day$total, main="Total Steps Per Day", xlab="Total Steps", ylab="Frequency")
# mean and median total steps taken per day
mean_steps <- mean(total_steps_per_day$total)
median_steps <- median(total_steps_per_day$total)
```  
The mean number of steps taken per day is `r format(mean_steps)` and the median number of steps taken per day is `r median_steps`

## What is the average daily activity pattern?

```{r averageDailyPattern, echo = T}
mean_steps_per_interval <- group_by(activity_data, interval) %>%
  filter(!is.na(steps)) %>%
  summarise(mean_steps=mean(steps))
with(mean_steps_per_interval, 
     plot(interval, mean_steps, type='l', main='Mean Steps per Interval', xlab="Time of Day", ylab="Mean Steps"))
most_steps <- with(mean_steps_per_interval, interval[which.max(mean_steps)])
```  

The 5-minute interval that contains the most steps, on average, is `r format(most_steps, "%H:%M")`

## Imputing missing values
```{r missingValues1, echo = T}
num_missing <- sum(is.na(activity_data$steps))
```
The number of missing values in the dataset is `r num_missing`.
```{r missingValues2, echo = T}
# If an interval has missing steps data, use the average number of steps for that interval for imputation
activity_data$steps[is.na(activity_data$steps)] <- sapply(activity_data$interval[is.na(activity_data$steps)],
                                                          function(x_interval) 
                                                            { with(mean_steps_per_interval,
                                                                   mean_steps[interval==x_interval])})
total_steps_per_day <- group_by(activity_data, date) %>%
  summarise(total=sum(steps))
# histogram of total steps
hist(total_steps_per_day$total, main="Total Steps Per Day", xlab="Total Steps", ylab="Frequency")
# mean and median total steps taken per day
mean_steps <- mean(total_steps_per_day$total)
median_steps <- median(total_steps_per_day$total)
```

The mean number of steps taken each day is `r format(mean_steps)`, which is the same mean as without imputed values. The median number of steps taken each day is `r format(median_steps)`, which has only shifted slightly from the median without imputed values and now equals the mean. Imputing missing values has very little impact on the estimates of the total daily number of steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdaysVsWeekends, echo = T}
isWeekend <- weekdays(activity_data$date, abbreviate=T) %in% c("Sat", "Sun")
activity_data$dayType <- factor(isWeekend, labels=c("weekday", "weekend"))
steps_by_interval_and_day <- group_by(activity_data, interval, dayType) %>%
  summarise(mean_steps=mean(steps))
library(ggplot2)
g <- ggplot(steps_by_interval_and_day, aes(x=interval,y=mean_steps,group=dayType)) +
  geom_line(aes(color=dayType)) +
  facet_grid(dayType~.) +
  xlab("Time of Day") +
  ylab("Mean Number of Steps") +
  ggtitle("Mean Steps by Time of Day (Weekday vs Weekend)")
g
```
